---
title: "IDE Spatial Mouse Liver Transriptomics"
subtitle: > 
  Initial Data Exploration Across Two Spatial Transcriptomics Seurat Objects
date: today
date-format: "DD.MM.YYYY"
author: 
  - name: Christian Kolland
    email: christian.kolland@stud.uni-frankfurt.de
    affiliation:
      - name: Goethe University Frankfurt
        city: Frankfurt am Main
abstract: >
  The IDE of mouse liver ST data generated using nanoString's CosMx™ technology. 
  The datasets are composed of single-cell gene expression counts from two 
  separate sample slides, which are stored in two separate Seurat objects.
copyright: 
  holder: Christian Kolland
  year: 2025
title-block-banner: "#1D2951"
toc: true
---

```{r}
#| label: load-packages
#| echo: false
#| include: false

box::use(
  dplyr[case_when, mutate, select],
  glue[glue],
  Seurat[AddMetaData],
  SeuratObject[LoadSeuratRds],
  tibble[as_tibble],
)
```

```{r}
#| label: load-utils
#| echo: false
#| include: false

box::use(
  tables = ./helpers/render_tables
)
```

## What Data Do We Have?

For this initial data exploration (IDE) we have spatial transcriptomics (ST)
data generated from mouse model liver tissue. Tissue samples were taken from
mice with different stages of alcoholic liver disease.

Following disease states are represented in the data:

1.  Healthy
2.  Steatosis
3.  Cirrhosis
4.  Acute-on-Chronic Liver Failure (ACLF)

### Data Generation

The data was generated using nanoString's CosMx™ spatial molecular imager (SMI).
This assay provides transcriptomics data at cellular and subcellular resolutions
using a non-damaging in situ approach. The samples were prepared and processed
on two separate slides. Each slide led to an individual [Seurat
object](https://satijalab.org/seurat/), which stores the resulting expression
counts and spatial information.

The [CosMx™ Mouse Neuroscience
Panel](https://nanostring.com/products/cosmx-spatial-molecular-imager/cosmx-rna-assays/cosmx-mouse-neuroscience-panel/) defined the set or genetic markers to be tracked. This leads to a certain
"brain" bias.

Markers that were used to visualize the cell morphology:

B2M
:   Beta-2 Microglobulin, a small protein that is a component of MHC class I
    molecules found on all nucleated cells. It is often used as a marker in
    immunohistochemistry.

CD298 (SLC12A2)
:   Protein associated with ion transport and is often used as a marker for
    certain cellular functions.

PanCK
:   Pan-Cytokeratin, a group of cytoskeletal proteins found in epithelial cells.
    It is used to identify epithelial cell populations in tissue.

CD45
:   Marker for leukocytes (white blood cells). It is used to identify immune
    cells in tissues.

CK8/18
:   Cytokeratin 8/18, which are intermediate filament proteins expressed in
    epithelial cells, particularly in simple epithelia such as those of the
    liver.

DAPI
:   4′,6-diamidino-2-phenylindole, a fluorescent stain that binds strongly to
    DNA, commonly used for nuclear staining in microscopy.

## Run Information

```{r}
#| label: run-information
#| echo: false

# Input directory
data_dir <- "../data/Seurat_objects"

# Output directories
run_dir <- glue("run_{format(Sys.time(), '%Y-%m-%d_%H-%M')}")
out_dir <- glue("../output/data/Seurat_objects/{run_dir}")

# Ensure the output directory exists
if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

message(glue("
📂 Input Seurat objects from: {data_dir}
💾 Outputs will be saved in: {out_dir}
"))
```

## Load the Data

```{r}
#| label: load-data-liver1

sobj_l1 <- LoadSeuratRds(file.path(
  data_dir,
  "Seurat_object_liver1.Rds"
))

sobj_l1
```

```{r}
#| label: load-data-liver2

sobj_l2 <- LoadSeuratRds(file.path(
  data_dir,
  "Seurat_object_liver2.Rds"
))

sobj_l2
```

## Data Exploration

::: {.callout-note title="Access Seurat Objects"}
To access the count information in a Seurat object:

```R
obj_l1@assays$Nanostring$counts
```

To access metadata in a Seurat object:

```R
sobj_l1@meta.data
```
:::

```{r}
#| label: access-counts

sobj_l1@assays$Nanostring$counts[1:5, 1:5]
```

Features (genes) are stored as rows in Seurat objects. 
Cell samples are stored in columns with their dataset-specific unique IDs. 

The count data is stored in a sparse matrix format.
&rarr; Only non-zero values are stored with their coordinates in the matrix

```{r}
#| label: access-metadata
#| tbl-cap: "Overview of available metadata (first five columns)"

sobj_l1@meta.data |>
  head(5) |>
  tables$render_table(data = _)
```

Metadata columns that will be important for this project will be:

1.  `Slide_name`
2.  `tissue`
3.  `fov`

`fov` assigns each cell to its corresponding tissue "stamp". A slide contains
multiple tissue slices, with each slice being "stamped" n times. The `fov` value
indicates in which stamp a given cell is located.

### Add Metadata `disease_state`

::: {.callout-note title="Disease State Metadata"}
The disease state must be added to the existing Seurat object, as this data is
not currently included in the `meta.data` table. The disease states are mapped
via the FOV IDs.

The disease state mapping can be found in the corresponding code blocks.
:::

```{r}
#| label: prepare-disease-state-liver1

# Disease state to FOV mapping
healthy <- c(1, 2, 15, 16, 17, 18, 19, 20, 21, 22, 23, 45)
cirrhosis <- c(5, 6, 24, 25, 26, 30, 31, 34, 35, 40, 41, 44)
steatosis <- c(3, 4, 9, 10, 13, 14, 27, 28, 29, 38, 39)
aclf <- c(7, 8, 11, 12, 32, 33, 36, 37, 42, 43)

# Add column "disease_state"
# Condition is assigned according to defined mapping
disease_states_l1 <- data.frame(
  fov = sobj_l1@meta.data$fov, # Conditions are mapped to FOVs
  row.names = colnames(sobj_l1)
) |>
  mutate(
    disease_state = case_when(
      fov %in% healthy ~ "healthy",
      fov %in% cirrhosis ~ "cirrhosis",
      fov %in% steatosis ~ "steatosis",
      fov %in% aclf ~ "aclf",
      TRUE ~ "other"
    )
  ) |>
  select(-fov)

unique(disease_states_l1) |>
  as_tibble()
```

```{r}
#| label: prepare-disease-state-liver2

# See "prepare-disease-state-liver1"
cirrhosis <- c(1, 2, 3, 4, 5, 6, 7, 8, 13, 14, 17, 18, 19, 20, 21, 22, 23, 24, 25, 36, 37, 44, 45)
aclf <- c(9, 10, 11, 12, 15, 16, 26, 27, 28, 29, 30, 31, 32, 33, 34, 35, 38, 39, 40, 41, 42, 43)

disease_states_l2 <- data.frame(
  fov = sobj_l2@meta.data$fov,
  row.names = colnames(sobj_l2)
) |>
  mutate(
    disease_state = case_when(
      fov %in% cirrhosis ~ "cirrhosis",
      fov %in% aclf ~ "aclf",
      TRUE ~ "other"
    )
  ) |>
  select(-fov)

unique(disease_states_l2) |>
  as_tibble()
```

**Add disease state metadata.**

```{r}
#| label: add-disease-state-liver1

# Add disease condition to metadata
sobj_l1 <- AddMetaData(
  sobj_l1,
  metadata = disease_states_l1
)
```

```{r}
#| label: check-conditions-liver1
#| echo: false

# Validate that correct disease states are added.
if (identical(
  unique(disease_states_l1)$disease_state,
  unique(sobj_l1@meta.data$disease_state)
)) {
  message("Metadata added correctly to Liver 1.")
} else {
  message("An error occured during metadata attachment...")
}
```

```{r}
#| label: add-conditions-liver2

sobj_l2 <- AddMetaData(
  sobj_l2,
  metadata = disease_states_l2
)
```

```{r}
#| label: check-conditions-liver2
#| echo: false

if (identical(
  unique(disease_states_l2)$disease_state,
  unique(sobj_l2@meta.data$disease_state)
)) {
  message("Metadata added correctly to Liver 2.")
} else {
  message("An error occured during metadata attachment...")
}
```

## Save the Data

```{r}
#| label: save-liver1

saveRDS(sobj_l1, file = file.path(out_dir, "Seurat_liver1.Rds"))
```

```{r}
#| label: save-liver2

saveRDS(sobj_l2, file = file.path(out_dir, "Seurat_liver2.Rds"))
```
