---
title: "EDA Spatial Mouse Liver Transriptomics"
subtitle: >
  Exploratory Data Analysis Across Spatial Transcriptomics Seurat Objects
date: today
date-format: "DD.MM.YYYY"
author: 
  - name: Christian Kolland
    email: christian.kolland@stud.uni-frankfurt.de
    affiliation:
      - name: Goethe University Frankfurt
        city: Frankfurt am Main
abstract: >
  The EDA of mouse liver ST data generated using nanoString's CosMx™ technology. 
  The datasets are composed of single-cell gene expression counts from two 
  separate sample slides, which are stored in two separate Seurat objects and
  merged to a single object.
copyright: 
  holder: Christian Kolland
  year: 2025
title-block-banner: "#1D2951"
toc: true
---

```{r}
#| label: load-packages
#| echo: false
#| include: false

box::use(
  arrow[write_feather],
  dplyr[bind_rows, case_when, count, filter, mutate, rename, select],
  ggplot2[ggsave],
  ggplot2,
  glue[glue],
  Matrix[rowSums],
  purrr[map],
  SeuratObject[LoadSeuratRds],
  tibble[as_tibble, tibble],
  tidyr[separate],
)
```

```{r}
#| label: load-utils
#| echo: false
#| include: false

box::use(
  plots = ./helpers/render_plots,
  tables = ./helpers/render_tables
)
```

## What Data Do We Have?

For this exploratory data analysis (EDA) we have the same spatial 
transcriptomics (ST) data generated from mouse model liver tissue that was 
initially explored in `IDE_mouse_liver_ST.qmd`.

## Run Information

```{r}
#| label: run-information
#| echo: false

# Input directories
data_dir <- "../data/Seurat_objects"

# Output directories
run_dir <- glue("run_{format(Sys.time(), '%Y-%m-%d_%H-%M')}")
out_dir <- glue("../output/EDA_mouse_liver_ST/{run_dir}")

# Ensure the output directory exists
if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

message(glue("
📂 Input Seurat objects from: {data_dir}
💾 Outputs will be saved under: {out_dir}
"))
```

## Load the Data

```{r}
#| label: load-liver1

sobj_l1 <- LoadSeuratRds(file.path(
  data_dir,
  "Seurat_liver1.Rds"
))

sobj_l1
```

```{r}
#| label: load-liver2

sobj_l2 <- LoadSeuratRds(file.path(
  data_dir,
  "Seurat_liver2.Rds"
))

sobj_l2
```

## Data Exploration

We aim to visualize a variety of features in order to better understand our ST 
datasets. Our goal is to explore the individual datasets in detail to assess 
their comparability and determine whether they can be meaningfully merged and 
integrated.

### Spatial Information

Plot all cells in a number of selected fields of view (FOV).

```{r}
#| label: fig-cell-map
#| fig-cap: "Spatial distribution of cells in liver1, FOVs 1 and 2."

# Plot cells in liver1 FOV 1 and 2
plot <- plots$render_cell_map(sobj_l1, fovs = c("1", "2"))

# Save as .png and .svg to be flexible
ggplot2$ggsave(file.path(out_dir, "cell_map_fov_1_2.png"))
ggplot2$ggsave(file.path(out_dir, "cell_map_fov_1_2.svg"))

plot
```

Observations @fig-cell-map:

- Not all parts of the FOV are filled with cells
  - Can lead to different cell count distributions across FOVs
  
Plot transcripts across cells in a selected FOV.

```{r}
#| label: fig-transcript-map
#| fig-cap: "Spatial expression of genes Pecam1, C1qc, and Dcn in liver1 FOV 1."

# Plot transcripts in liver1 FOV 1
plot <- plots$render_transcript_map(
  sobj_l1,
  fovs = c("1"),
  genes = c("Pecam1", "C1qc", "Dcn")
)

ggplot2$ggsave(file.path(out_dir, "transcript_map_fov_1.png"))
ggplot2$ggsave(file.path(out_dir, "transcript_map_fov_1.svg"))

plot
```

Observations @fig-transcript-map:

- Exact locations were transcripts are measured
  - Marker genes can visually quantify cell types

### Counts

This section examines the individual count distributions within both datasets.

#### `nCount_Nanostring`

Value found in the `meta.data` table in our Seurat objects.

```{r}
#| label: tbl-validate-nCount-Nanostring
#| tbl-cap: |
#|   Validation that nCount_Nanostring in metadata equals the sum of raw counts 
#|   per cell.

# Names of first five cells in liver1 dataset
# Extracted from metadata table (cells are columns in Seurat)
row_names_first_5 <- rownames(sobj_l1@meta.data)[1:5]

# Counts for first five cells
counts_first_5 <- sobj_l1@assays$Nanostring$counts[, row_names_first_5]

# Sum of counts for first five cells
count_sums_first_5 <- map(row_names_first_5, ~ sum(counts_first_5[, .x]))

# Create tibble for plotting
tibble(
  "sum" = count_sums_first_5,
  "nCounts_Nanostring" = sobj_l1@meta.data$nCount_Nanostring[1:5]
) |>
  tables$render_table(data = _)
```

In @tbl-validate-nCount-Nanostring we see that `nCount_Nanostring` is equal to sum 
of counts for each individual cell.

##### Descriptive Statistics

```{r}
#| label: tbl-descriptive-statistics-cell-counts
#| tbl-cap: |
#|   Descriptive statistics for total counts per cell in liver1 and liver2 
#|   datasets.

# Calculate descriptive stats for counts per cell
cell_counts_stats_l1 <- tibble(
  "scope" = "cell",
  "dataset" = "liver1",
  "min" = min(sobj_l1@meta.data$nCount_Nanostring),
  "max" = max(sobj_l1@meta.data$nCount_Nanostring),
  "median" = median(sobj_l1@meta.data$nCount_Nanostring),
  "mean" = mean(sobj_l1@meta.data$nCount_Nanostring),
)

count_counts_stats_l2 <- tibble(
  "scope" = "cell",
  "dataset" = "liver2",
  "min" = min(sobj_l2@meta.data$nCount_Nanostring),
  "max" = max(sobj_l2@meta.data$nCount_Nanostring),
  "median" = median(sobj_l2@meta.data$nCount_Nanostring),
  "mean" = mean(sobj_l2@meta.data$nCount_Nanostring),
)

# Combine tibbles for plotting
# Combine different count statistics in this tibble
desc_stats_counts <- cell_counts_stats_l1 |>
  bind_rows(count_counts_stats_l2)

# Plot counts per cell statistics
desc_stats_counts |>
  filter(scope == "cell") |> # Counts per cell
  select(-scope) |>
  tables$render_table(data = _)
```

##### Plots

```{r}
#| label: fig-cell-counts-density
#| fig-cap: |
#|   Kernel density estimates of total counts per cell in both datasets.

# Prepare data for plotting
# Sum of counts per cell
cell_counts_dist_l1 <- as_tibble(sobj_l1@meta.data$nCount_Nanostring) |>
  mutate(dataset = "liver1")

cell_counts_dist_l2 <- as_tibble(sobj_l2@meta.data$nCount_Nanostring) |>
  mutate(dataset = "liver2")

# Combine to plotting data
# Save plotting data later as it is changed
pdata_counts_per_cell <- cell_counts_dist_l1 |>
  dplyr::bind_rows(cell_counts_dist_l2)

plot <- plots$render_density_plot(
  pdata_counts_per_cell,
  dataset,
  legend = "Datasets",
  title = "Count Distributions: Liver1 vs. Liver2",
  x_lab = "Counts per Cell"
)

ggplot2$ggsave(file.path(out_dir, "density_counts_per_cell.png"))
ggplot2$ggsave(file.path(out_dir, "density_counts_per_cell.svg"))

plot
```

Observations @fig-cell-counts-density:

- Both curves are positive (mean greater then median)
- Mean appears to be around a total expression of ~500
- Tail on right signals bigger amount of outlier cells with very high expression
- Looks like negative binomial distribution
  - Similar to scRNA-seq counts
- Similar distributions &rarr; **enables dataset integration**

```{r}
#| label: fig-cell-counts-violin
#| fig-cap: "Violin plot comparing per-cell counts across datasets."

plot <- plots$render_violin_plot(
  pdata_counts_per_cell,
  dataset,
  legend = "Datasets",
  title = "Count Distributions: Liver1 vs. Liver2",
  x_lab = "Datasets",
  y_lab = "Counts per Cell"
)

ggplot2$ggsave(file.path(out_dir, "violin_counts_per_cell.png"))
ggplot2$ggsave(file.path(out_dir, "violin_counts_per_cell.svg"))

plot
```

Observations @fig-cell-counts-violin:

- median value for both distributions
  - Value at ~800 counts per cell
- Similar distribution for outliers
  - liver2 has some outliers that have counts above 7000
- **Similarity between distributions is reconfirmed**
  
```{r}
#| label: fig-cell-counts-ECDF
#| fig-cap: "ECDF plot showing cumulative distribution of counts per cell."

plot <- plots$render_ecdf_plot(
  pdata_counts_per_cell,
  dataset,
  legend = "Datasets",
  title = "Count Distributions: Liver1 vs. Liver2",
  x_lab = "Counts per Cell",
  y_lab = "Cumulative Probability"
)

ggplot2$ggsave(file.path(out_dir, "ECDF_counts_per_cell.png"))
ggplot2$ggsave(file.path(out_dir, "ECDF_counts_per_cell.svg"))

plot
```

Observations @fig-cell-counts-ECDF:

- Around 50% of cells have summarized counts less then 1000
- Around 88% of cells have summarized counts less then 2000
- Small amount of outliers with very high expressions

#### `nFeature_Nanostring`

Value found in the `meta.data` table in our Seurat objects.

```{r}
#| label: tbl-validate-nFeature-Nanostring
#| tbl-cap: |
#|   Validation that nFeature_Nanostring in metadata equals the sum of non-zero 
#|   features per cell.

# Sum all non-zero features in a cell
feature_sums_first_5 <- map(row_names_first_5, ~ sum(counts_first_5[, .x] != 0))

# Create tibble for plotting
tibble(
  "sum" = count_sums_first_5,
  "nFeature_Nanostring" = sobj_l1@meta.data$nFeature_Nanostring[1:5]
) |>
  tables$render_table(data = _)
```

In @tbl-validate-nFeature-Nanostring we see that `nFeature_Nanostring` is equal to 
sum of of all non-zero features for each individual cell.

##### Descriptive Statistics

```{r}
#| label: tbl-descriptive-statistics-feature-counts
#| tbl-cap: |
#|   Descriptive statistics for non-zero features (genes) detected per cell 
#|   across datasets.

# Calculate descriptive stats non-zero features per cell
cell_feats_stats_l1 <- tibble(
  "scope" = "feature",
  "dataset" = "liver1",
  "min" = min(sobj_l1@meta.data$nFeature_Nanostring),
  "max" = max(sobj_l1@meta.data$nFeature_Nanostring),
  "median" = median(sobj_l1@meta.data$nFeature_Nanostring),
  "mean" = mean(sobj_l1@meta.data$nFeature_Nanostring),
)

cell_feats_stats_l2 <- tibble(
  "scope" = "feature",
  "dataset" = "liver2",
  "min" = min(sobj_l2@meta.data$nFeature_Nanostring),
  "max" = max(sobj_l2@meta.data$nFeature_Nanostring),
  "median" = median(sobj_l2@meta.data$nFeature_Nanostring),
  "mean" = mean(sobj_l2@meta.data$nFeature_Nanostring),
)

# Add scope to count statistics
desc_stats_counts <- desc_stats_counts |>
  bind_rows(cell_feats_stats_l1) |>
  bind_rows(cell_feats_stats_l2)

# Plot non-zero features per cell statistics
desc_stats_counts |>
  filter(scope == "feature") |>
  select(-scope) |>
  tables$render_table(data = _)
```

##### Plots

```{r}
#| label: fig-cell-features-density
#| fig-cap: "Density plot of detected non-zero features (genes) per cell."

# Prepare data for plotting
# Sum of features per cell
cell_feats_dist_l1 <- as_tibble(sobj_l1@meta.data$nFeature_Nanostring) |>
  mutate(dataset = "liver1")

cell_feats_dist_l2 <- as_tibble(sobj_l2@meta.data$nFeature_Nanostring) |>
  mutate(dataset = "liver2")

pdata_features_per_cell <- cell_feats_dist_l1 |>
  bind_rows(cell_feats_dist_l2)

# Save plotting data
# .feather good for data transfer between R and Python
write_feather(
  pdata_features_per_cell,
  glue("{out_dir}/pdata_features_per_cell.feather")
)

plot <- plots$render_density_plot(
  pdata_features_per_cell,
  dataset,
  legend = "Datasets",
  title = "Non-zero Feature Distributions: Liver1 vs. Liver2",
  x_lab = "Non-zero Features per Cell"
)

ggplot2$ggsave(file.path(out_dir, "density_features_per_cell.png"))
ggplot2$ggsave(file.path(out_dir, "density_features_per_cell.svg"))

plot
```

Observations @fig-cell-features-density:

- Almost normal distributions
  - Small tail on the right
- No unexpected results
- Similar distributions &rarr; **enables dataset integration**

#### Counts per Gene

```{r}
#| label: calculate-gene-counts

# Calculate sum of counts per gene for both datasets
# Matrix::rowSums for sparse matrices
gene_counts_l1 <- rowSums(sobj_l1@assays$Nanostring$counts) |>
  as_tibble() |>
  mutate("dataset" = "liver1")

gene_counts_l2 <- rowSums(sobj_l2@assays$Nanostring$counts) |>
  as_tibble() |>
  mutate("dataset" = "liver2")
```

##### Descriptive Statistics

```{r}
#| label: tbl-descriptive-statistics-gene-counts
#| tbl-cap: |
#|   Descriptive statistics for gene-level total counts in liver1 and liver2.

# Calculate descriptive stats counts per gene
gene_counts_stats_l1 <- tibble(
  "scope" = "gene",
  "dataset" = "liver1",
  "min" = min(gene_counts_l1$value),
  "max" = max(gene_counts_l1$value),
  "median" = median(gene_counts_l1$value),
  "mean" = mean(gene_counts_l1$value),
)

gene_counts_stats_l2 <- tibble(
  "scope" = "gene",
  "dataset" = "liver2",
  "min" = min(gene_counts_l2$value),
  "max" = max(gene_counts_l2$value),
  "median" = median(gene_counts_l2$value),
  "mean" = mean(gene_counts_l2$value),
)

# Add scope to count statistics
desc_stats_counts <- desc_stats_counts |>
  bind_rows(gene_counts_stats_l1) |>
  bind_rows(gene_counts_stats_l2)

# Plot counts per gene statistics
desc_stats_counts |>
  filter(scope == "gene") |>
  select(-scope) |>
  tables$render_table(data = _)
```

##### Plots

```{r}
#| label: fig-gene-counts-density
#| fig-cap: |
#|   "Log-transformed density plot of total counts per gene across both datasets."

# Combine data for plotting
pdata_counts_per_gene <- gene_counts_l1 |>
  bind_rows(gene_counts_l2)

# Preprocess for better plotting
pdata_counts_per_gene <- pdata_counts_per_gene |> 
  rename(raw_counts = value) |> 
  mutate(value = log10(raw_counts + 1))

write_feather(
  pdata_counts_per_gene,
  glue("{out_dir}/pdata_counts_per_gene.feather")
)

plot <- plots$render_density_plot(
  pdata_counts_per_gene,
  dataset,
  legend = "Datasets",
  title = "Count Distributions: Liver1 vs. Liver2",
  x_lab = "log10(Counts per Gene + 1)"
)

ggplot2$ggsave(file.path(out_dir, "density_counts_per_gene.png"))
ggplot2$ggsave(file.path(out_dir, "density_counts_per_gene.svg"))

plot
```

Observations @fig-gene-counts-density:

- Values are log10 normalized
  - Values near 0 mean no expression
  - Higher values indicate higher expression
- Compared to scRNA-seq data we see high average expression
- No 0 expressed genes (see @descriptive-statistics-gene-counts)
- Similar distributions &rarr; **enables dataset integration**

### `fov`

```{r}
#| label: explore-fov-liver1

unique(sobj_l1@meta.data$fov)
```

```{r}
#| label: explore-fov-liver2

unique(sobj_l2@meta.data$fov)
```

Equal number of FOVs. Need to be properly split in merged dataset.

:::{.callout-important title="Number of Unique FOVs"}
Need to be 90 FOVs in merged dataset.
:::

### Visualize UMAPs

Plot the precalculated UMAPs from both datasets in one UMAP.

```{r}
#| label: fig-precalculated-umap-slide
#| fig-cap: "UMAP visualization colored by slide origin."

# Extract UMAP coordinates from both datasets
# Also add metadata for plotting
precalc_umap_l1 <- tibble(
  "dataset" = "liver1",
  "umap1" = sobj_l1@reductions$umap@cell.embeddings[, 1],
  "umap2" = sobj_l1@reductions$umap@cell.embeddings[, 2],
  "slide" = sobj_l1@meta.data$Slide_name,
  "disease_state" = sobj_l1@meta.data$disease_state,
  "cell_type" = sobj_l1@meta.data$nb_clus
)

precalc_umap_l2 <- tibble(
  "dataset" = "liver2",
  "umap1" = sobj_l2@reductions$umap@cell.embeddings[, 1],
  "umap2" = sobj_l2@reductions$umap@cell.embeddings[, 2],
  "slide" = sobj_l2@meta.data$Slide_name,
  "disease_state" = sobj_l2@meta.data$disease_state,
  "cell_type" = sobj_l2@meta.data$nb_clus
)

pdata_precalculated_umap <- precalc_umap_l1 |>
  bind_rows(precalc_umap_l2)

write_feather(
  pdata_precalculated_umap,
  glue("{out_dir}/pdata_precalculated_umap.feather")
)

plot <- plots$render_umap_plot(
  pdata_precalculated_umap,
  slide,
  legend =  "Slide",
  title = "Precalculated UMAP"
)

ggplot2$ggsave(file.path(out_dir, "precalc_umap_slide.png"))
ggplot2$ggsave(file.path(out_dir, "precalc_umap_slide.svg"))

plot
```

Observations @fig-precalculated-umap-slide:

- Both samples show same UMAP shape
- There appears to be a different batch effect

```{r}
#| label: fig-precalc-umap-disease-state
#| fig-cap: "UMAP visualization colored by cell disease state."

plot <- plots$render_umap_plot(
  pdata_precalculated_umap,
  disease_state,
  legend =  "Disease State",
  title = "Precalculated UMAP"
)

ggplot2$ggsave(file.path(out_dir, "precalc_umap_disease_state.png"))
ggplot2$ggsave(file.path(out_dir, "precalc_umap_disease_state.svg"))

plot
```

Observations @fig-precalc-umap-disease-state:

- Clusters according to disease state
  - ACLF cells on top and bottom cluster
- Another stronger batch effect

```{r}
#| label: fig-precalc-umap-cell-type
#| fig-cap: "UMAP visualization colored by nanoString annotated cell types."

plot <- plots$render_umap_plot(
  pdata_precalculated_umap,
  cell_type,
  legend =  "Cell Type",
  title = "Precalculated UMAP"
)

ggplot2$ggsave(file.path(out_dir, "precalc_umap_cell_type.png"))
ggplot2$ggsave(file.path(out_dir, "precalc_umap_cell_type.svg"))

plot
```

Observations @fig-precalc-umap-cell-type:

- Cell type does not appear to be batch effect
  - **Batch effect needs to be removed**

## Load Merged Data

```{r}
#| label: load-merged-liver

sobj_liver_merged <- LoadSeuratRds(file.path(
  data_dir,
  "Seurat_liver_merged.Rds"
))

sobj_liver_merged
```

### `fov`

```{r}
#| label: check-fovs

# Gather all unique FOVs
merged_fovs <- unique(sobj_liver_merged@meta.data$fov)
# Number of unique FOVs
n_fovs <- length(merged_fovs)

n_fovs
```

::: {.callout-tip title="FOVs in Merged Dataset"}
90 unique FOVs in the merged dataset.
&rarr; As expected
:::

### Plots

```{r}
#| label: fig-density-disease-state
#| fig-cap: "Density plot of counts per cell grouped by disease state."

# Counts per cell with disease state information
pdata_cell_counts_per_x <- pdata_counts_per_cell |>
  mutate(disease_state = sobj_liver_merged@meta.data$disease_state)

write_feather(
  pdata_cell_counts_per_x,
  file.path(out_dir, "pdata_counts_per_cell.feather")
)

plot <- plots$render_density_plot(
  pdata_cell_counts_per_x,
  disease_state,
  legend = "Disease State",
  title = "Count Distributions Across Disease States",
  x_lab = "Counts per Cell"
)

ggplot2$ggsave(file.path(out_dir, "density_cell_counts_per_disease_state.png"))
ggplot2$ggsave(file.path(out_dir, "density_cell_counts_per_disease_state.svg"))

plot
```

Observations @fig-density-disease-state:

- Healthy very positively skewed
  - Highest peak
- Similar mean expression over all states
- Other three conditions with more variability
- All have outliers

```{r}
#| label: fig-density-cells-per-fov
#| fig-cap: "Horizontal bar plot of cell counts per FOV, grouped by dataset."

# Create cells per FOV distribution data
pdata_cells_per_fov <- tibble(fov = sobj_liver_merged@meta.data$fov) |>
  count(fov, name = "value") |>
  separate(col = fov, into = c("group", "dataset"), sep = "_")

write_feather(
  pdata_cells_per_fov,
  file.path(out_dir, "pdata_cells_per_fov.feather")
)

plot <- plots$render_bar_plot(
  pdata_cells_per_fov,
  dataset,
  legend = "Datasets",
  title = "Cells per FOV: Liver1 vs. Liver2",
  x_lab = "Number of Cells",
  y_lab = "FOV"
)

ggplot2$ggsave(file.path(out_dir, "bars_cells_per_fov.png"))
ggplot2$ggsave(file.path(out_dir, "bars_cells_per_fov.svg"))

plot
```

Observations @fig-density-cells-per-fov:

- Reveals heterogeneous count distributions
  - Cells are not similar distributed across FOVs in the dataset
  - Cells are not similar distributed across FOVs over both datasets
- Expected, since tissue samples are "random"
  - Not 100% of the FOV is tissue
  - Big gaps lead to less cells
