---
title: "EDA of Mouse Liver Spatial Transcriptomics Data"
date: today
date-format: "DD.MM.YYYY"
author: 
  - name: Christian Kolland
    email: christian.kolland@stud.uni-frankfurt.de
    affiliation:
      - name: Goethe University Frankfurt
        city: Frankfurt am Main
abstract: >
  Exploratory Data Analysis (EDA) of spatial transcriptomics data generated using 
  NanoString's CosMx™ technology. The data set consists of single cell gene 
  expression counts from two separate slides of tissue samples derived from 
  mice representing four different disease states: healthy, steatosis, cirrhosis 
  and Acute-on-Chronic Liver Failure (ACLF).
copyright: 
  holder: Christian Kolland
  year: 2025
title-block-banner: "#002147"
---

```{r}
#| label: load-packages
#| echo: false
#| include: false

box::use(
  dplyr[case_when, count, mutate, select],
  gt[gt],
  ggplot2,
  glue[glue],
  patchwork[plot_annotation],
  Seurat[AddMetaData, DimPlot, RunPCA, RunUMAP, ScaleData],
  SeuratObject[LoadSeuratRds],
  tibble[as_tibble],
)
```

```{r}
#| label: load-utils
#| echo: false
#| include: false

box::use(
  ./utils
)
```

## Run Information

```{r}
#| label: load-params
#| echo: false

# Input directories
data_dir <- "../data"
sobj_dir <- file.path(data_dir, "seurat_objects")

# Output directories
out_dir <- "../output"
eda_out <- file.path(out_dir, glue("eda_output_{format(Sys.time(), '%d-%m-%Y_%H-%M')}"))

message(glue("
📂 Input Seurat objects from: {sobj_dir}
💾 Outputs will be saved under: {eda_out}
"))
```

## What Data Do We Have?

The data examined in this EDA consists of mouse model data related to alcoholic 
liver disease. The data comprises two slides that were prepared with a number of 
Formalin-Fixed Paraffin-Embedded (FFPE) liver samples from different disease 
states.

Following disease states are represented in the data:

1. Healthy
2. Steatosis
3. Cirrhosis
4. ACLF

The data was generated by two different slides (stored in serrated Seurat
objects).

Markers used to visualize the cell morphology:

B2M
: Beta-2 Microglobulin, a small protein that is a component of MHC class I 
molecules found on all nucleated cells. It is often used as a marker in 
immunohistochemistry.

CD298 (SLC12A2)
: Protein associated with ion transport and is often used as a marker for 
certain cellular functions.

PanCK
: Pan-Cytokeratin, a group of cytoskeletal proteins found in epithelial cells. 
It is used to identify epithelial cell populations in tissue.

CD45
: Marker for leukocytes (white blood cells). It is used to identify immune cells 
in tissues.

CK8/18
: Cytokeratin 8/18, which are intermediate filament proteins expressed in 
epithelial cells, particularly in simple epithelia such as those of the liver.

DAPI
: 4′,6-diamidino-2-phenylindole, a fluorescent stain that binds strongly to DNA, 
commonly used for nuclear staining in microscopy.

## Load the Data

```{r}
#| label: load-data-liver1

sobj_liver1 <- LoadSeuratRds(file.path(
  data_dir,
  "seurat_objects/seurat_object_liver1.Rds"
))
sobj_liver1
```

```{r}
#| label: load-data-liver2

sobj_liver2 <- LoadSeuratRds(file.path(
  data_dir,
  "seurat_objects/seurat_object_liver2.Rds"
))
sobj_liver2
```

## Data Exploration

:::{.callout-note title="Access Seurat Object"}
To access the count information in the Seurat object:
```
sobj_liver1@assays$Nanostring$counts
```

To access metadata:
```
sobj_liver1@meta.data
```
:::

### Counts

```{r}
#| label: explore-counts-l1-1

sobj_liver1@assays$Nanostring$counts[1:5, 1:5]
```

:::{.callout-note title="Count Matrix"}
The count matrix is setup as a sparse matrix.
:::

### Metadata

```{r}
#| label: explore-metadata
#| echo: false

sobj_liver1@meta.data |>
  head(10) |>
  gt()
```

Same for both Seurat object: liver1 and liver2.

#### `nCount_Nanostring`

Sum of counted features inside the cell.

```{r}
#| label: sum-counts-row-1

row_name_sample <- rownames(sobj_liver1@meta.data)[sobj_liver1@meta.data$id == "liver1_3_1"]
counts_sample <- sobj_liver1@assays$Nanostring$counts[, row_name_sample]
count_sum_sample <- sum(counts_sample)
```

| `nCount_Nanostring` | Summed Counts |
|---|---|
| `r sobj_liver1@meta.data$nCount_Nanostring[sobj_liver1@meta.data$id == "liver1_3_1"]`| `r count_sum_sample` |

: Sanity check to validate metadata value

##### Statistical Parameters

```{r}
#| label: statistical-ncount
#| echo: false

min_val <- min(sobj_liver1@meta.data$nCount_Nanostring)
max_val <- max(sobj_liver1@meta.data$nCount_Nanostring)
mean_val <- round(mean(sobj_liver1@meta.data$nCount_Nanostring), 4)

message(glue("
📊 nCount_Nanostring @ sobj_liver1
🔹 Min: {min_val}
🔹 Max: {max_val}
🔹 Mean: {mean_val}
"))
```

```{r}
#| label: plot-ncount
#| echo: false

# Plotting parameters
bin_size <- 50

# Plot liver1
p_count_distribution_l1 <- as_tibble(sobj_liver1@meta.data$nCount_Nanostring) |>
  ggplot2$ggplot(ggplot2$aes(x = value)) +
  ggplot2$geom_histogram(bins = bin_size, fill = "#a6cee3", color = "#1f78b4") +
  ggplot2$labs(
    x = "Counts per Cell",
    y = "Counts",
    title = glue("Liver1 (Bin Size: {bin_size})")
  ) +
  # Should be applied before theme
  ggplot2$theme_bw() + # White background
  ggplot2$theme_minimal() + # No background annotation
  ggplot2$theme(
    plot.title = ggplot2$element_text(size = 11)
  )

# Plot liver2
p_count_distribution_l2 <- as_tibble(sobj_liver2@meta.data$nCount_Nanostring) |>
  ggplot2$ggplot(ggplot2$aes(x = value)) +
  ggplot2$geom_histogram(bins = bin_size, fill = "#a6cee3", color = "#1f78b4") +
  ggplot2$labs(
    x = "Counts per Cell",
    y = "Counts",
    title = glue("Liver2 (Bin Size: {bin_size})")
  ) +
  # Should be applied before theme
  ggplot2$theme_bw() + # White background
  ggplot2$theme_minimal() + # No background annotation
  ggplot2$theme(
    plot.title = ggplot2$element_text(size = 11)
  )

# Find the maximum y value from both histograms
y_limit <- max(
  max(ggplot2$ggplot_build(p_count_distribution_l1)$data[[1]]$count),
  max(ggplot2$ggplot_build(p_count_distribution_l2)$data[[1]]$count)
)

# Adjust y-axis limits to be the same for both histograms
# Makes them comparable
p_count_distribution_l1 <- p_count_distribution_l1 + ggplot2$ylim(0, y_limit)
p_count_distribution_l2 <- p_count_distribution_l2 + ggplot2$ylim(0, y_limit)

# Merge plots to one plot
comp_count_distributions <- p_count_distribution_l1 + p_count_distribution_l2 +
  plot_annotation(title = "Per Cell Count Distribution")


ggplot2$ggsave(
  file.path(eda_out, "fig_per_cell_count_distribution.png"),
  comp_count_distributions,
  create.dir = TRUE
)
message(glue("💾  Plot saved as: fig_per_cell_count_distribution"))

comp_count_distributions
```

#### `nFeature_Nanostring`

Sum of all non-zero features found in the cell.

```{r}
#| label: sum-features-row-1

feature_sum_sample <- sum(counts_sample != 0)
```

| `nFeature_Nanostring` | Non-zero Features |
|---|---|
| `r sobj_liver1@meta.data$nFeature_Nanostring[sobj_liver1@meta.data$id == "liver1_3_1"]`| `r feature_sum_sample` |

: Sanity check to validate metadata value

##### Statistical Parameters

```{r}
#| label: statistical-nfeature
#| echo: false

min_val <- min(sobj_liver1@meta.data$nFeature_Nanostring)
max_val <- max(sobj_liver1@meta.data$nFeature_Nanostring)
mean_val <- mean(sobj_liver1@meta.data$nFeature_Nanostring)

message(glue("
📊 nFeature_Nanostring @ sobj_liver1
🔹 Min: {min_val}
🔹 Max: {max_val}
🔹 Mean: {mean_val}
"))
```

```{r}
#| label: plot-nfeature
#| echo: false

# Plotting parameters
bin_size <- 50

# Plot liver1
p_feature_distribution_l1 <- as_tibble(sobj_liver1@meta.data$nFeature_Nanostring) |>
  ggplot2$ggplot(ggplot2$aes(x = value)) +
  ggplot2$geom_histogram(bins = bin_size, fill = "#b2df8a", color = "#33a02c") +
  ggplot2$labs(
    x = "Non-zero Features per Cell",
    y = "Counts",
    title = glue("Liver1 (Bin Size: {bin_size})")
  ) +
  # Should be applied before theme
  ggplot2$theme_bw() + # White background
  ggplot2$theme_minimal() + # No background annotation
  ggplot2$theme(
    plot.title = ggplot2$element_text(size = 11)
  )

# Plot liver2
p_feature_distribution_l2 <- as_tibble(sobj_liver1@meta.data$nFeature_Nanostring) |>
  ggplot2$ggplot(ggplot2$aes(x = value)) +
  ggplot2$geom_histogram(bins = bin_size, fill = "#b2df8a", color = "#33a02c") +
  ggplot2$labs(
    x = "Non-zero Features per Cell",
    y = "Counts",
    title = glue("Liver2 (Bin Size: {bin_size})")
  ) +
  # Should be applied before theme
  ggplot2$theme_bw() + # White background
  ggplot2$theme_minimal() + # No background annotation
  ggplot2$theme(
    plot.title = ggplot2$element_text(size = 11)
  )

# Find the maximum y value from both histograms
y_limit <- max(
  max(ggplot2$ggplot_build(p_feature_distribution_l1)$data[[1]]$count),
  max(ggplot2$ggplot_build(p_feature_distribution_l2)$data[[1]]$count)
)

# Adjust y-axis limits to be the same for both histograms
# Makes them comparable
p_feature_distribution_l1 <- p_feature_distribution_l1 + ggplot2$ylim(0, y_limit)
p_feature_distribution_l2 <- p_feature_distribution_l2 + ggplot2$ylim(0, y_limit)

# Merge plots to one plot
comp_feature_distributions <- p_feature_distribution_l1 + p_feature_distribution_l2 +
  plot_annotation(title = "Per Cell Feature Distribution")


ggplot2$ggsave(
  file.path(eda_out, "fig_per_cell_feature_distribution.png"),
  comp_feature_distributions,
  create.dir = TRUE
)
message(glue("💾  Plot saved as: fig_per_cell_feature_distribution"))

comp_feature_distributions
```

#### `fov`

Assigns each cell to its corresponding tissue "stamp".
A slide contains multiple tissue slices, with each slice being "stamped" n times.
The `fov` value indicates on which stamp a given cell is located.

```{r}
#| label: explore-fov-l1

unique(sobj_liver1@meta.data$fov)
```

```{r}
#| label: explore-fov-l2

unique(sobj_liver2@meta.data$fov)
```

```{r}
#| label: plot-fov
#| echo: false

# Prepare plotting data
fov_distribution_l1 <- sobj_liver1@meta.data$fov |>
  as_tibble() |>
  count(value, sort = TRUE)

fov_distribution_l2 <- sobj_liver2@meta.data$fov |>
  as_tibble() |>
  count(value, sort = TRUE)

# Plotting parameters
y_limit <- max(fov_distribution_l1$n, fov_distribution_l2$n) # Determine y ceiling

# Plot liver1
p_fov_distribution_l1 <- fov_distribution_l1 |>
  ggplot2$ggplot(ggplot2$aes(x = factor(value, levels = unique(value)), y = n)) +
  ggplot2$geom_col(fill = "#6a3d9a") +
  ggplot2$ylim(0, y_limit) +
  ggplot2$scale_x_discrete(guide = ggplot2$guide_axis(n.dodge = 2)) + # Staggers x-axis ticks
  ggplot2$labs(
    x = "FOVs",
    y = "Counted Cells",
    title = "Liver1"
  ) +
  # Should be applied before theme
  ggplot2$theme_bw() + # White background
  ggplot2$theme_minimal() + # No background annotation
  ggplot2$theme(
    axis.text.x = ggplot2$element_text(angle = 270, hjust = 1),
    plot.title = ggplot2$element_text(size = 11)
  )

# Plot liver2
p_fov_distribution_l2 <- fov_distribution_l2 |>
  ggplot2$ggplot(ggplot2$aes(x = factor(value, levels = unique(value)), y = n)) +
  ggplot2$geom_col(fill = "#6a3d9a") +
  ggplot2$ylim(0, y_limit) +
  ggplot2$scale_x_discrete(guide = ggplot2$guide_axis(n.dodge = 2)) + # Staggers x-axis ticks
  ggplot2$labs(
    x = "FOVs",
    y = "Counted Cells",
    title = "Liver2"
  ) +
  # Should be applied before theme
  ggplot2$theme_bw() + # White background
  ggplot2$theme_minimal() + # No background annotation
  ggplot2$theme(
    axis.text.x = ggplot2$element_text(angle = 270, hjust = 1),
    plot.title = ggplot2$element_text(size = 11)
  )

# Merge plots to one plot
comp_fov_distributions <- p_fov_distribution_l1 + p_fov_distribution_l2 +
  plot_annotation(title = "Cell Distribution Over FOVs")

ggplot2$ggsave(
  file.path(eda_out, "fig_cell_distribution_fovs.png"),
  comp_fov_distributions,
  create.dir = TRUE
)
message(glue("💾  Plot saved as: fig_cell_distribution_fovs"))

comp_fov_distributions
```

#### `condition`

```{r}
#| label: disease-conditions-l1

# Condition to FOV mapping
healthy <- c(1, 2, 15, 16, 17, 18, 19, 20, 21, 22, 23, 45)
cirrhosis <- c(5, 6, 24, 25, 26, 30, 31, 34, 35, 40, 41, 44)
steatosis <- c(3, 4, 9, 10, 13, 14, 27, 28, 29, 38, 39)
aclf <- c(7, 8, 11, 12, 32, 33, 36, 37, 42, 43)

# Add column "condition"
# Condition is assigned according to defined mapping
disease_conditions_l1 <- data.frame(
  fov = sobj_liver1@meta.data$fov,
  row.names = colnames(sobj_liver1)
) |>
  mutate(
    condition = case_when(
      fov %in% healthy ~ "healthy",
      fov %in% cirrhosis ~ "cirrhosis",
      fov %in% steatosis ~ "steatosis",
      fov %in% aclf ~ "aclf",
      TRUE ~ "Other"
    )
  ) |>
  select(-fov)

unique(disease_conditions_l1)
```

:::{.callout-important title="Expected Values Liver1"}
1. healthy
2. steatosis
3. cirrhosis
4. aclf
:::

```{r}
#| label: disease-conditions-l2

# Condition to FOV mapping
cirrhosis <- c(1, 2, 3, 4, 5, 6, 7, 8, 13, 14, 17, 18, 19, 20, 21, 22, 23, 24, 25, 36, 37, 44, 45)
aclf <- c(9, 10, 11, 12, 15, 16, 26, 27, 28, 29, 30, 31, 32, 33, 34, 35, 38, 39, 40, 41, 42, 43)

disease_conditions_l2 <- data.frame(
  fov = sobj_liver2@meta.data$fov,
  row.names = colnames(sobj_liver2)
) |>
  mutate(
    condition = case_when(
      fov %in% cirrhosis ~ "cirrhosis",
      fov %in% aclf ~ "aclf",
      TRUE ~ "Other"
    )
  ) |>
  select(-fov)

unique(disease_conditions_l2)
```

:::{.callout-important title="Expected Values Liver2"}
1. cirrhosis
2. aclf
:::

```{r}
#| label: prepare-plotting-condition

## Prepare data for plotting
# Add disease condition to metadata
sobj_liver1 <- AddMetaData(
  sobj_liver1,
  metadata = disease_conditions_l1
)

# Count conditions over samples to get distribution
condition_distribution_l1 <- sobj_liver1@meta.data$condition |>
  as_tibble() |>
  count(value, sort = TRUE)

sobj_liver2 <- AddMetaData(
  sobj_liver2,
  metadata = disease_conditions_l2
)

# Count conditions over samples to get distribution
condition_distribution_l2 <- sobj_liver2@meta.data$condition |>
  as_tibble() |>
  count(value, sort = TRUE)
```

```{r}
#| label: plot-condition
#| echo: false

# Plotting parameters
y_limit <- max(condition_distribution_l1$n, condition_distribution_l2$n) # Determine y ceiling

# Plot liver1
p_condition_distribution_l1 <- condition_distribution_l1 |>
  ggplot2$ggplot(ggplot2$aes(x = factor(value, levels = unique(value)), y = n)) +
  ggplot2$geom_col(fill = "#cab2d6") +
  ggplot2$ylim(0, y_limit) +
  ggplot2$labs(
    x = "Conditions",
    y = "Counted Cells",
    title = "Liver1"
  ) +
  # Should be applied before theme
  ggplot2$theme_bw() + # White background
  ggplot2$theme_minimal() + # No background annotation
  ggplot2$theme(
    axis.text.x = ggplot2$element_text(angle = 45, hjust = 1),
    plot.title = ggplot2$element_text(size = 11)
  )

# Plot liver2
p_condition_distribution_l2 <- condition_distribution_l2 |>
  ggplot2$ggplot(ggplot2$aes(x = factor(value, levels = unique(value)), y = n)) +
  ggplot2$geom_col(fill = "#cab2d6") +
  ggplot2$ylim(0, y_limit) +
  ggplot2$labs(
    x = "Conditions",
    y = "Counted Cells",
    title = "Liver2"
  ) +
  # Should be applied before theme
  ggplot2$theme_bw() + # White background
  ggplot2$theme_minimal() + # No background annotation
  ggplot2$theme(
    axis.text.x = ggplot2$element_text(angle = 45, hjust = 1),
    plot.title = ggplot2$element_text(size = 11)
  )

# Merge plots to one plot
comp_condition_distributions <- p_condition_distribution_l1 + p_condition_distribution_l2 +
  plot_annotation(title = "Cell Distribution Over Conditions")

ggplot2$ggsave(
  file.path(eda_out, "fig_cell_distribution_conditions.png"),
  comp_condition_distributions,
  create.dir = TRUE
)
message(glue("💾  Plot saved as: fig_cell_distribution_conditions"))

comp_condition_distributions
```

## (Optional) Merge Objects

Only merges objects if saved version is not available.

Needs to increase memory limit to process.
`options(future.globals.maxSize = 3000 * 1024^2)  # Sets limit to ~4,5GB`

```{r}
#| label: merge-objects

if (!file.exists(file.path(data_dir, "seurat_objects/sobj_merged_liver.Rds"))) {
  # Sets limit to ~4,5GB to enable proper computation for large Seurat objects
  options(future.globals.maxSize = 4500 * 1024^2)

  sobj_liver_merged <- utils$merge_seurat_objects(
    sobj_liver1,
    sobj_liver2,
    ids = c("liver1", "liver2"),
    save_dir = sobj_dir
  )
} else {
  sobj_liver_merged <- readRDS(file.path(sobj_dir, "sobj_merged_liver.Rds"))
}

sobj_liver_merged
```

:::{.callout-important title="Expected Values Merged Liver"}
34452 + 33333 = 67785 samples
:::

## Visualize UMAPs

```{r}
#| label: prepare-plotting-umap

## Create object to plot combined UMAP data
# Object for liver1
liver1_reduced <- as.data.frame(t(as.matrix(sobj_liver1@assays$Nanostring@counts)))
liver1_reduced$Cell_Type <- sobj_liver1@meta.data$nb_clus
liver1_reduced$Slide <- sobj_liver1@meta.data$Slide_name
liver1_reduced$Condition <- sobj_liver1@meta.data$condition

# Add UMAP coordinates
liver1_reduced$umap1 <- sobj_liver1@reductions$umap@cell.embeddings[, 1]
liver1_reduced$umap2 <- sobj_liver1@reductions$umap@cell.embeddings[, 2]

# Object for liver2
liver2_reduced <- as.data.frame(t(as.matrix(sobj_liver2@assays$Nanostring@counts)))
liver2_reduced$Cell_Type <- sobj_liver2@meta.data$nb_clus
liver2_reduced$Slide <- sobj_liver2@meta.data$Slide_name
liver2_reduced$Condition <- sobj_liver2@meta.data$condition

# Add UMAP coordinates
liver2_reduced$umap1 <- sobj_liver2@reductions$umap@cell.embeddings[, 1]
liver2_reduced$umap2 <- sobj_liver2@reductions$umap@cell.embeddings[, 2]

# Merge objects
liver_reduced <- as.data.frame(rbind(liver1_reduced, liver2_reduced))
liver_reduced |>
  head(10) |> 
  gt()
```

```{r}
#| label: plot-umap-condition
#| echo: false

p_umap_condition <- liver_reduced |>
  ggplot2$ggplot(ggplot2$aes(x = umap1, y = umap2, col = Condition)) +
  ggplot2$geom_point() +
  ggplot2$theme_bw() +
  ggplot2$theme_minimal()

ggplot2$ggsave(
  file.path(eda_out, "fig_umap_condition.png"),
  p_umap_condition,
  create.dir = TRUE
)
message(glue("💾  Plot saved as: fig_umap_condition"))

p_umap_condition
```

```{r}
#| label: plot-umap-slide
#| echo: false

p_umap_slide <- liver_reduced |>
  ggplot2$ggplot(ggplot2$aes(x = umap1, y = umap2, col = Slide)) +
  ggplot2$geom_point() +
  ggplot2$theme_bw() +
  ggplot2$theme_minimal()

ggplot2$ggsave(
  file.path(eda_out, "fig_umap_slide.png"),
  p_umap_slide,
  create.dir = TRUE
)
message(glue("💾  Plot saved as: fig_umap_slide"))

p_umap_slide
```
