---
title: Exploratory Data Analysis
author: Christian Kolland
date: today
---

```{r}
#| label: load-packages
#| echo: false
#| include: false

box::use(
  dplyr,
  DT,
  ggplot2,
  patchwork,
  Seurat,
  SeuratObject,
  tibble
)
```

```{r}
#| label: load-utils
#| echo: false
#| include: false

#' @export
get_border_boundaries <- function(sobj, fovs) {
  box::use(
    Seurat[Images]
  )

  # Images() returns the slide name --> Only one value
  slide_name <- Images(sobj)[1]
  # Returns the first tissue name in a vector
  # Apparently all the same value
  tissue_name <- sobj@meta.data$Run_Tissue_name[1]

  # `id` returns the cell id
  # `fov` must be part of the given FOVs (only take chosen ones)
  # Tissue must be identical?
  # Returns all cells that are part of the FOVs
  cells_in_boundary <- sobj$id[
    (sobj$fov %in% fovs) &
      (sobj$Run_Tissue_name == tissue_name)
  ]

  # Returns all centroids for the slide
  centroids <- sobj@images[[slide_name]]$centroids
  # Renames all cells according to the slide
  centroids@cells <- paste(slide_name, centroids@cells, sep = "_")

  # Returns 4 coordinates to build a boundary box for the cells of interest
  border_boundaries <- apply(
    centroids@coords[centroids@cells %in% cells_in_boundary, ],
    2,
    range
  )

  border_boundaries
}

#' @export
plot_cells_in_fov <- function(sobj, fovs) {
  box::use(
    ggplot2[xlim, ylim],
    Seurat[ImageDimPlot, Images]
  )

  # Get the border boundaries for ours FOVs
  fov_boundaries <- get_border_boundaries(sobj, fovs)
  slide_name <- Images(sobj)[1]

  ImageDimPlot(
    sobj,
    fov = slide_name,
    border.color = "black"
  ) +
    xlim(fov_boundaries[, 2]) +
    ylim(fov_boundaries[, 1])
}

#' @export
plot_gene_expression_markers <- function(sobj, fovs, genes) {
  box::use(
    ggplot2[xlim, ylim],
    Seurat[ImageDimPlot]
  )

  fov_boundaries <- get_border_boundaries(sobj, fovs)
  slide_name <- Images(sobj)[1]

  ImageDimPlot(
    sobj,
    fov = slide_name,
    border.color = "black",
    alpha = 0.3, # Reduce alpha of cell fills to improve molecule visualization
    molecules = genes,
    mols.size = 0.8,
    nmols = 100000, # Set the total number of molecules to visualize
    axes = FALSE
  ) +
    xlim(fov_boundaries[, 2]) +
    ylim(fov_boundaries[, 1])
}
```

## What Data do We Have?

The data analyzed in this EDA consists of mouse model data related to alcoholic 
liver disease. The data consists of two FFPE samples (slides).

:::{.callout-note title="FFPE"}
The acronym FFPE stands for "formalin-fixed paraffin-embedded".
:::

The samples stem from the CosMx Mouse Neuroscience Panel (product of 
[nanoString](https://nanostring.com/products/cosmx-spatial-molecular-imager/cosmx-rna-assays/cosmx-mouse-neuroscience-panel/)) 
Enables for expression profiling of 1000 highly curated targets at subcellular 
resolution. Customizable with up to 50 of own targets.

Morphology visualization for (these are markers):

B2M
: Beta-2 Microglobulin, a small protein that is a component of MHC class I 
molecules found on all nucleated cells. It is often used as a marker in 
immunohistochemistry.

CD298 (SLC12A2)
: Protein associated with ion transport and is often used as a marker for 
certain cellular functions.

PanCK
: Pan-Cytokeratin, a group of cytoskeletal proteins found in epithelial cells. 
It is used to identify epithelial cell populations in tissue.

CD45
: Marker for leukocytes (white blood cells). It is used to identify immune cells 
in tissues.

CK8/18
: Cytokeratin 8/18, which are intermediate filament proteins expressed in 
epithelial cells, particularly in simple epithelia such as those of the liver.

DAPI
: 4′,6-diamidino-2-phenylindole, a fluorescent stain that binds strongly to DNA, 
commonly used for nuclear staining in microscopy.

## Load the Data

The Seurat objects contain 4 different types of information:

1. Read counts per gene per cell
2. Location of cells on slide
3. Inferred information (cell cluster, cell type)
4. Cell annotation (with respect to field of view (FOV))

```{r}
#| label: load-data-liver1

sobj_liver1 <- SeuratObject$LoadSeuratRds("../data/liver1_seurat_object.Rds")
sobj_liver1
```

```{r}
#| label: load-data-liver2

sobj_liver2 <- SeuratObject$LoadSeuratRds("../data/liver2_seurat_object.Rds")
sobj_liver2
```

## Data Exploration

:::{.callout-note title="Access Seurat object"}
To access the count information in the Seurat object:

```
sobj_liver1@assays$Nanostring$counts
```

To access other metadata:

```
sobj_liver1@meta.data
```
:::

```{r}
#| label: explore-metadata-1

sobj_liver1@meta.data
```

```{r}
#| label: explore-metadata-2

colnames(sobj_liver1@meta.data)
```

Each column holds the metadata of a specific type for all 
`r nrow(sobj_liver1@meta.data)` samples.

### nCount_Nanostring

Apparently holds the number of counted features inside the cell.
This means the first cell had **450** counted features inside of it.

```{r}
#| label: explore-metadata-ncount-1

sobj_liver1@meta.data$nCount_Nanostring[1:10]
```

```{r}
#| label: explore-metadata-ncount-2

min_val <- min(sobj_liver1@meta.data$nCount_Nanostring)
max_val <- max(sobj_liver1@meta.data$nCount_Nanostring)
mean_val <- mean(sobj_liver1@meta.data$nCount_Nanostring)

message(paste0("Min: ", min_val, ", Max: ", max_val, ", Mean: ", mean_val))
```

```{r}
#| label: plot-metadata-ncount

p_count_dist_l1 <- tibble$as_tibble(sobj_liver1@meta.data$nCount_Nanostring) |>
  ggplot2$ggplot(ggplot2$aes(x = value)) +
  ggplot2$geom_histogram(bins = 20, fill = "#a6cee3", color = "#1f78b4") +
  ggplot2$ylim(0, 10000) +
  ggplot2$labs(
    x = "Values",
    y = "Counts"
  ) +
  ggplot2$theme_minimal()

p_count_dist_l2 <- tibble$as_tibble(sobj_liver2@meta.data$nCount_Nanostring) |>
  ggplot2$ggplot(ggplot2$aes(x = value)) +
  ggplot2$geom_histogram(bins = 20, fill = "#a6cee3", color = "#1f78b4") +
  ggplot2$ylim(0, 10000) +
  ggplot2$labs(
    x = "Values",
    y = "Counts"
  ) +
  ggplot2$theme_minimal()

comb_count_plot <- p_count_dist_l1 + p_count_dist_l2 +
  patchwork$plot_annotation(
    title = "Count Distributions",
    caption = "Liver1 vs. Liver2"
  )

ggplot2$ggsave("../output/fig_count_distribution.png", comb_count_plot)
comb_count_plot
```

### nFeature_Nanostring

Number of individual features found in one cell.
The first cell had **181** individual features counted inside of it.

```{r}
#| label: explore-metadata-nfeature-1

sobj_liver1@meta.data$nFeature_Nanostring[1:10]
```

```{r}
#| label: explore-metadata-nfeature-2

min_val <- min(sobj_liver1@meta.data$nFeature_Nanostring)
max_val <- max(sobj_liver1@meta.data$nFeature_Nanostring)
mean_val <- mean(sobj_liver1@meta.data$nFeature_Nanostring)

message(paste0("Min: ", min_val, ", Max: ", max_val, ", Mean: ", mean_val))
```

```{r}
#| label: plot-metadata-nfeature

p_feat_dist_l1 <- tibble$as_tibble(sobj_liver1@meta.data$nFeature_Nanostring) |>
  ggplot2$ggplot(ggplot2$aes(x = value)) +
  ggplot2$geom_histogram(bins = 20, fill = "#b2df8a", color = "#33a02c") +
  ggplot2$labs(
    x = "Values",
    y = "Counts"
  ) +
  ggplot2$theme_minimal()

p_feat_dist_l2 <- tibble$as_tibble(sobj_liver2@meta.data$nFeature_Nanostring) |>
  ggplot2$ggplot(ggplot2$aes(x = value)) +
  ggplot2$geom_histogram(bins = 20, fill = "#b2df8a", color = "#33a02c") +
  ggplot2$labs(
    x = "Values",
    y = "Counts"
  ) +
  ggplot2$theme_minimal()

comb_feat_plot <- p_feat_dist_l1 + p_feat_dist_l2 +
  patchwork$plot_annotation(
    title = "Feature Distributions",
    caption = "Liver1 vs. Liver2"
  )

ggplot2$ggsave("../output/fig_feature_distribution.png", comb_feat_plot)
comb_feat_plot
```

### fov

Provides all tissue stamps of a slide.
A slide contains a number of tissue slices. Each tissue slice is "stamped" n 
times. `fov` tells on which stamp the given cell can be found.

```{r}
#| label: explore-metadata-fov-liver1

unique(sobj_liver1@meta.data$fov)
```

```{r}
#| label: plot-metadata-fov-liver1

fov_dist_liver1 <- sobj_liver1@meta.data$fov |>
  tibble$as_tibble() |>
  dplyr$count(value, sort = TRUE)

p_fov_dist_l1 <- fov_dist_liver1 |>
  ggplot2$ggplot(ggplot2$aes(x = as.factor(value), y = n)) +
  ggplot2$geom_col(fill = "#cab2d6") +
  ggplot2$ylim(0, 1250) +
  ggplot2$labs(
    x = "FOVs",
    y = "Counts"
  ) +
  ggplot2$theme_minimal() +
  ggplot2$theme(
    axis.text.x = ggplot2$element_text(angle = 90, hjust = 1)
  )
```

```{r}
#| label: explore-metadata-fov-liver2

unique(sobj_liver2@meta.data$fov)
```

```{r}
#| label: plot-metadata-fov-liver2

fov_dist_liver2 <- sobj_liver2@meta.data$fov |>
  tibble$as_tibble() |>
  dplyr$count(value, sort = TRUE)

p_fov_dist_l2 <- fov_dist_liver2 |>
  ggplot2$ggplot(ggplot2$aes(x = as.factor(value), y = n)) +
  ggplot2$geom_col(fill = "#cab2d6") +
  ggplot2$ylim(0, 1250) +
  ggplot2$labs(
    x = "FOVs",
    y = "Counts"
  ) +
  ggplot2$theme_minimal() +
  ggplot2$theme(
    axis.text.x = ggplot2$element_text(angle = 90, hjust = 1)
  )
```

```{r}
#| label: plot-metadata-fov-combined

comb_fov_plot <- p_fov_dist_l1 + p_fov_dist_l2 +
  patchwork$plot_annotation(
    title = "Sample Counts per FOV",
    caption = "Liver1 vs. Liver2"
  )

ggplot2$ggsave("../output/fig_fov_distribution.png", comb_fov_plot)
comb_fov_plot
```

### Counts

```{r}
#| label: explore-counts-l1-1

sobj_liver1@assays$Nanostring$counts[1:5, 1:5]
```

```{r}
#| label: explore-counts-l1-2

num_rows <- nrow(sobj_liver1@assays$Nanostring$counts)
num_cols <- ncol(sobj_liver1@assays$Nanostring$counts)

message(paste0("Rows: ", num_rows, ", Cols: ", num_cols))
```

A gene to cell matrix (usually genes on cols).

```{r}
#| label: explore-counts-l2

num_rows <- nrow(sobj_liver2@assays$Nanostring$counts)
num_cols <- ncol(sobj_liver2@assays$Nanostring$counts)

message(paste0("Rows: ", num_rows, ", Cols: ", num_cols))
```

## Visualization

```{r}
#| label: plot-cells-fov-1-2-l1

p_cells_fov_1_2_l1 <- plot_cells_in_fov(sobj_liver1, c(1, 2))

ggplot2$ggsave("../output/fig_cells_fov_1_2_l1.png", p_cells_fov_1_2_l1)
p_cells_fov_1_2_l1
```

```{r}
#| label: plot-cells-fov-1-2-l2

p_cells_fov_1_2_l2 <- plot_cells_in_fov(sobj_liver2, c(1, 2))

ggplot2$ggsave("../output/fig_cells_fov_1_2_l2.png", p_cells_fov_1_2_l2)
p_cells_fov_1_2_l2
```

```{r}
#| label: plot-gexprm-fov-1-2-l1

p_gexprm_fov_1_2_l1 <- plot_gene_expression_markers(
  sobj_liver1,
  c(1, 2),
  c("Slc17a7", "Rgs5")
)

ggplot2$ggsave("../output/fig_gexprm_fov_1_2_l1.png", p_gexprm_fov_1_2_l1)
p_gexprm_fov_1_2_l1
```

```{r}
#| label: plot-feat-rgs5-l1

p_feat_rgs5_l1 <- Seurat$FeaturePlot(
  sobj_liver1,
  features = "Rgs5",
  order = TRUE
)

ggplot2$ggsave("../output/fig_feature_rgs5_l1.png", p_feat_rgs5_l1)
p_feat_rgs5_l1
```

## Disease Conditions

```{r}
#| label: disease-conditions-l1

healthy <- c(1, 2, 15, 16, 17, 18, 19, 20, 21, 22, 23, 45)
cirrhosis <- c(5, 6, 24, 25, 26, 30, 31, 34, 35, 40, 41, 44)
steatosis <- c(3, 4, 9, 10, 13, 14, 27, 28, 29, 38, 39)
aclf <- c(7, 8, 11, 12, 32, 33, 36, 37, 42, 43)

conditions_l1 <- sobj_liver1@meta.data$fov |>
  tibble$as_tibble() |>
  dplyr$rename(FOV = value) |>
  # dplyr$distinct() |>
  dplyr$mutate(
    condition = dplyr$case_when(
      FOV %in% healthy ~ "healthy",
      FOV %in% cirrhosis ~ "cirrhosis",
      FOV %in% steatosis ~ "steatosis",
      FOV %in% aclf ~ "aclf",
      TRUE ~ "Other"
    )
  )

conditions_l1
```

```{r}
#| label: disease-conditions-l2

cirrhosis <- c(1, 2, 3, 4, 5, 6, 7, 8, 13, 14, 17, 18, 19, 20, 21, 22, 23, 24, 25, 36, 37, 44, 45)
aclf <- c(9, 10, 11, 12, 15, 16, 26, 27, 28, 29, 30, 31, 32, 33, 34, 35, 38, 39, 40, 41, 42, 43)

conditions_l2 <- sobj_liver2@meta.data$fov |>
  tibble$as_tibble() |>
  dplyr$rename(FOV = value) |>
  # dplyr$distinct() |>
  dplyr$mutate(
    condition = dplyr$case_when(
      FOV %in% cirrhosis ~ "cirrhosis",
      FOV %in% aclf ~ "aclf",
      TRUE ~ "Other"
    )
  )

conditions_l2
```

## Prepare Combined Object

All three operations can not be run due to memory constraints.
Need to be outsourced to cluster.

```{r}
#| label: add-metadata-l1
#| eval: false

sobj_liver1 <- Seurat$AddMetaData(
  sobj_liver1,
  metadata = conditions_l1 |>
    dplyr$select(-FOV)
)
```

```{r}
#| label: add-metadata-l2
#| eval: false

sobj_liver2 <- Seurat$AddMetaData(
  sobj_liver2,
  metadata = conditions_l2 |>
    dplyr$select(-FOV)
)
```

Rename cells with slide prefix.

```{r}
#| label: merge-objects

sobj_combined_liver <- merge(sobj_liver1, sobj_liver2)
```

