---
title: "Seurat to AnnData Conversion"
subtitle: "Extraction of relevant data from the Seurat object"
date: today
date-format: "DD.MM.YYYY"
author: 
  - name: Christian Kolland
    email: christian.kolland@stud.uni-frankfurt.de
    affiliation:
      - name: Goethe University Frankfurt
        city: Frankfurt am Main
abstract: >
  Extract the relevant data from the merged liver Seurat object to prepare it 
  for conversion into an AnnData object in Python. This includes the expression 
  matrix as well as metadata such as condition, field of view (FOV), and other 
  relevant annotations.
copyright: 
  holder: Christian Kolland
  year: 2025
title-block-banner: "#1D2951"
---

```{r}
#| label: load-packages
#| echo: false
#| include: false

box::use(
  arrow[write_feather],
  dplyr[mutate, select],
  glue[glue],
  Matrix[writeMM],
  readr[write_csv],
  SeuratObject[LoadSeuratRds],
  tibble[tibble]
)
```

## Run Information

```{r}
#| label: run-information
#| echo: false

# Input directory
data_dir <- "../data/Seurat_objects"

# Output directories
run_dir <- glue("run_{format(Sys.time(), '%Y-%m-%d_%H-%M')}")
out_dir <- glue("../output/data/Seurat_to_h5ad/{run_dir}")

# Ensure the output directory exists
if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

message(glue("
📂 Input Seurat objects from: {data_dir}
💾 Outputs will be saved in: {out_dir}
"))
```

## Load the Data

```{r}
#| label: load-data

sobj_liver_merged <- LoadSeuratRds(file.path(
  data_dir,
  "Seurat_liver_merged.Rds"
))

sobj_liver_merged
```

## Counts

To efficiently store the counts in their sparse representation, they are saved 
in the **Matrix Market (MM) format**. The data is stored in an `.mtx` file using the 
coordinate format, where only non-zero values are stored along with their 
respective row and column indices in the matrix.

For more details on the **Matrix Market format**, refer to the 
[official documentation](https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=5bce84be62e12ffe4d8a63fd118e4cd42f512807).

```{r}
#| label: extract-counts

# Save sparse counts
counts <- sobj_liver_merged@assays$Nanostring$counts

# Write to format that keeps sparsity in tact
write_out <- writeMM(counts, file.path(out_dir, "Seurat_counts.mtx"))

# Print for visual confirmation
counts |> head(10)
```

### Sanity Checks

```{r}
#| label: sanity-checks-counts

sanity_checks <- tibble(
  n_rows = nrow(counts), # Matrix rows
  n_cols = ncol(counts), # Matrix columns
  count_sum = sum(counts), # Sum of counts
  spotcheck_1 = list(as.matrix(counts[1:5, 1:5])), # Random count spots
  spotcheck_2 = list(as.matrix(counts[483:487, 33890:33894])),
  spotcheck_3 = list(as.matrix(counts[966:970, 67781:67785]))
)
```

## Extract Row and Column Names

Row and column names are not stored inside the `.mtx` file created above. 
Therefor this information needs to be stored separately.

### Row Names

```{r}
#| label: extract-row-names

# Save row names to csv
write_csv(
  as.data.frame(rownames(counts)),
  file.path(out_dir, "Seurat_row_names.csv"),
  col_names = FALSE
)

as.data.frame(rownames(counts)) |> head(10)
```

### Column Names

```{r}
#| label: extract-col-names
#| 
# Save column names to csv
write_csv(
  as.data.frame(colnames(counts)),
  file.path(out_dir, "Seurat_col_names.csv"),
  col_names = FALSE
)

as.data.frame(colnames(counts)) |> head(10)
```

### Sanity Checks

```{r}
#| label: sanity-checks-rows-cols

sanity_checks <- sanity_checks |>
  mutate(
    row_names = list(rownames(counts)),
    col_names = list(colnames(counts))
  )
```

## Extract Metadata

Identify if these values are identical.

```{r}
#| label: compare-tissue-RunTissueName

tissues <- sobj_liver_merged@meta.data$tissue
run_tissues <- sobj_liver_merged@meta.data$Run_Tissue_name
identical(tissues, run_tissues)
```

```{r}
#| label: extract-metadata

# Store metadata as
metadata <- sobj_liver_merged@meta.data |>
  select(
    nCount_Nanostring,
    nFeature_Nanostring,
    cell_ID,
    disease_state,
    Slide_name,
    tissue,
    fov
  )

# Write to specified output directory
write_csv(metadata, file.path(out_dir, "Seurat_metadata.csv"))

metadata |> head(10)
```

### Saniy Checks

```{r}
#| label: sanity-checks-metadata

sanity_checks <- sanity_checks |> 
  mutate(
    n_fovs = length(unique(metadata$fov)), # Must be 90 (45 for each slide)
    spotcheck_4 = list(metadata[1:5,]),
    spotcheck_5 = list(metadata[33890:33894,]),
    spotcheck_6 = list(metadata[67781:67785,])
  )

# Store sanity checks as .feather to make them usable in Python
write_feather(sanity_checks, file.path(out_dir, "sanity_checks.feather"))
```
