---
title: "IDE Spatial Mouse Liver Transriptomics"
date: today
date-format: "DD.MM.YYYY"
author: 
  - name: Christian Kolland
    email: christian.kolland@stud.uni-frankfurt.de
    affiliation:
      - name: Goethe University Frankfurt
        city: Frankfurt am Main
abstract: >
  Initial Data Exploration (IDE) of mouse liver spatial transcriptomics (ST) 
  data generated using NanoString's CosMx™ technology. The data set consists of 
  single-cell gene expression counts from two separate slides of tissue samples 
  stored in two separate Seurat objects.
copyright: 
  holder: Christian Kolland
  year: 2025
title-block-banner: "#002147"
---

```{r}
#| label: load-packages
#| echo: false
#| include: false

box::use(
  dplyr[case_when, mutate, select],
  gt[gt],
  glue[glue],
  Seurat[AddMetaData],
  SeuratObject[LoadSeuratRds],
  tibble[as_tibble],
)
```

## Run Information

```{r}
#| label: run-information
#| echo: false

# Input directory
data_dir <- "../data/Seurat_objects"

# Output directories
run_dir <- glue("run_{format(Sys.time(), '%Y-%m-%d_%H-%M')}")
out_dir <- glue("../output/Seurat_objects/{run_dir}")

# Ensure the output directory exists
if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

message(glue("
📂 Input Seurat objects from: {data_dir}
💾 Outputs will be saved in: {out_dir}
"))
```

## What Data Do We Have?

The data examined in this IDA consists of mouse model data related to alcoholic 
liver disease. The data comprises two slides that were prepared with a number of 
liver samples from different disease states. A [Seurat object](https://satijalab.org/seurat/) 
was created for each respective slide.

Following disease states are represented in the data:

1. Healthy
2. Steatosis
3. Cirrhosis
4. ACLF

:::{.callout-note title="Disease State Metadata"}
The disease state must be added to the existing Seurat object, as this data is 
not currently included in the `meta.data` table. The disease states are mapped 
via the FOV IDs.

The disease state mapping can be found in the corresponding code blocks.
:::

## Load the Data

```{r}
#| label: load-data-liver1

sobj_liver1 <- LoadSeuratRds(file.path(
  data_dir,
  "Seurat_object_liver1.Rds"
))
sobj_liver1
```

```{r}
#| label: load-data-liver2

sobj_liver2 <- LoadSeuratRds(file.path(
  data_dir,
  "Seurat_object_liver2.Rds"
))
sobj_liver2
```

## Data Exploration

:::{.callout-note title="Access Seurat Object"}
To access the count information in the Seurat object:
```
sobj_liver1@assays$Nanostring$counts
```

To access metadata:
```
sobj_liver1@meta.data
```
:::

Features (genes) are stored as rows in Seurat objects. 
Cell samples are stored in columns with their dataset-specific unique IDs.
The count data is stored in sparse matrix format.

```{r}
#| label: access-counts

sobj_liver1@assays$Nanostring$counts[1:5, 1:5]
```

Metadata columns that will be important for this project will be:

1. `Slide_name`
2. `tissue`
3. `fov`

`fov` assigns each cell to its corresponding tissue "stamp".
A slide contains multiple tissue slices, with each slice being "stamped" n times.
The `fov` value indicates in which stamp a given cell is located.

```{r}
#| label: access-metadata

sobj_liver1@meta.data |> 
  head(5) |> 
  gt()
```

### Add Metadata `disease_state`

```{r}
#| label: prepare-disease-states-l1

# Disease state to FOV mapping
healthy <- c(1, 2, 15, 16, 17, 18, 19, 20, 21, 22, 23, 45)
cirrhosis <- c(5, 6, 24, 25, 26, 30, 31, 34, 35, 40, 41, 44)
steatosis <- c(3, 4, 9, 10, 13, 14, 27, 28, 29, 38, 39)
aclf <- c(7, 8, 11, 12, 32, 33, 36, 37, 42, 43)

# Add column "disease_state"
# Condition is assigned according to defined mapping
disease_states_l1 <- data.frame(
  fov = sobj_liver1@meta.data$fov, # Conditions are mapped to FOVs
  row.names = colnames(sobj_liver1)
) |>
  mutate(
    disease_state = case_when(
      fov %in% healthy ~ "healthy",
      fov %in% cirrhosis ~ "cirrhosis",
      fov %in% steatosis ~ "steatosis",
      fov %in% aclf ~ "aclf",
      TRUE ~ "Other"
    )
  ) |>
  select(-fov)

unique(disease_states_l1) |> 
  as_tibble()
```

:::{.callout-important title="Expected Values Liver1"}
1. healthy
2. steatosis
3. cirrhosis
4. aclf
:::

```{r}
#| label: prepare-disease-states-l2

cirrhosis <- c(1, 2, 3, 4, 5, 6, 7, 8, 13, 14, 17, 18, 19, 20, 21, 22, 23, 24, 25, 36, 37, 44, 45)
aclf <- c(9, 10, 11, 12, 15, 16, 26, 27, 28, 29, 30, 31, 32, 33, 34, 35, 38, 39, 40, 41, 42, 43)

disease_states_l2 <- data.frame(
  fov = sobj_liver2@meta.data$fov,
  row.names = colnames(sobj_liver2)
) |>
  mutate(
    disease_state = case_when(
      fov %in% cirrhosis ~ "cirrhosis",
      fov %in% aclf ~ "aclf",
      TRUE ~ "Other"
    )
  ) |>
  select(-fov)

unique(disease_states_l2) |> 
  as_tibble()
```

:::{.callout-important title="Expected Values Liver2"}
1. cirrhosis
2. aclf
:::

Add disease state metadata to the Seurat objects.

```{r}
#| label: add-disease-states-l1

# Add disease condition to metadata
sobj_liver1 <- AddMetaData(
  sobj_liver1,
  metadata = disease_states_l1
)

if (identical(unique(disease_states_l1)$disease_state, unique(sobj_liver1@meta.data$disease_state))) {
  message("Metadata added correctly.")
} else {
  message("An error occured during metadata attachment...")
}
```

```{r}
#| label: add-conditions-l2

sobj_liver2 <- AddMetaData(
  sobj_liver2,
  metadata = disease_states_l2
)

if (identical(unique(disease_states_l2)$disease_state, unique(sobj_liver2@meta.data$disease_state))) {
  message("Metadata added correctly.")
} else {
  message("An error occured during metadata attachment...")
}
```

## Save the Data

```{r}
#| label: save-object-l1

saveRDS(sobj_liver1, file = file.path(out_dir, "sobj_liver1.Rds"))
```

```{r}
#| label: save-object-l2

saveRDS(sobj_liver2, file = file.path(out_dir, "sobj_liver2.Rds"))
```
